Learning pleiotropic signatures to construct a causal map of the human phenome

### 2017-07-03 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Towards the construction of a causal map of the human phenome},
            pdfauthor={Gibran Hemani, Jie Zheng, Philip Haycock, Oliver Davies, Jack Bowden, Tom Gaunt, George Davey Smith},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Towards the construction of a causal map of the human phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Gibran Hemani, Jie Zheng, Philip Haycock, Oliver Davies, Jack Bowden,
Tom Gaunt, George Davey Smith}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{03 July 2017}

@@

Mendelian randomisation (MR) (1,2) exploits genetic pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy. Such a genetic variant is considered to be an instrumental
variable for the exposure, and can be exploited to mimic a randomised
controlled trial, enabling a causal estimate to be made by comparing the
outcome phenotypes between those individuals that have the
exposure-increasing allele against those who do not. Multiple
independent genetic variants for a particular exposure can be used
jointly to improve causal inference, because a) each variant represents
an independent natural experiment, and an overall causal estimate can be
obtained by meta-analysing the single estimates from each instrument;
and b) some violations of the assumptions of MR (most notably that the
SNP influences the outcome through no pathways other than the exposure)
can be evaluated through sensitivity analyses (3--7).

Genome-wide association studies (GWAS) have identified genetic
instrumental variables for thousands of phenotypes (8). Recent
developments in Mendelian randomisation have enabled knowledge of
instrumental variables to be applied using only summary level data
(known as two-sample MR, 2SMR) (9). Here, in order to infer the causal
effect of an exposure on an outcome all that is required is an estimate
of the genetic effects of the instrumenting SNP on the exposure, and the
corresponding estimate of the effect on the outcome. This has two major
advantages. First, GWAS summary data is non-disclosive and often
publicly available. Second, causal inference can be made between
phenotypes even if they have not been measured in the same samples,
limiting the breadth of possible causal estimates only to the
availability of GWAS summary data for the traits in question (10).

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure. The
extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of HP (3--6). It is considered best
practice to report estimates from all available methods as sensitivity
analyses when presenting causal estimates, however this strategy is not
necessarily optimal for several reasons. First, if different methods
disagree it is not possible to know which is correct because the true
nature of HP exhibited by the instruments is not known. Second, though
the IVW approach is most statistically powerful under no HP, it can have
high false negative or low true positive rates in the presence of HP
compared to other methods. Given that pleiotropy has been hypothesised
to be universal (11,12), defaulting to the IVW method in the first
instance and using other methods as sensitivity analyses may not be
appropriate. Third, the available methods do not cover all possible
models of HP, and therefore an automated method for instrument selection
may be necessary. Fourth, it could be of interest to make causal effect
estimates for thousands of traits, in which case a discerning evaluation
of each causal effect of interest may not be possible or convenient.

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we introduce an approach to discard
genetic variants that are likely to be invalid. Second, we hypothesised
that characteristics of the summary data could indicate which method
would be most reliable, and we introduce new machine learning approaches
that attempt to automate both instrument and method selection. Using
curated GWAS summary data for thousands of phenotypes (10), we use these
new methods to construct a graph of millions of causal estimates. We
consider this to be a `first draft' of the causal map of the human
phenome, but raise caution throughout that its interpretation is far
from straightforward

@@

The use of summary data in two-sample MR is described in detail
elsewhere (10). A brief outline of the procedure is as follows. First,
genetic instruments for the exposure trait need to be identified - those
SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
assumption of Mendelian randomisation (that the instrument associates
with the exposure) is satisfied. We collect their effect sizes, standard
errors and effect alleles for the association with the exposure trait.
These can be obtanied manually from complete summary data or from
curated lists of significant GWAS associations. Next, the effects of
those SNPs on the outcome need to be obtained, typically necessitating
complete summary data because it is unlikely that these SNPs will have
reached genome-wide significance (and therefore be present in curated
catalogues). At its most simple implementation, the regression of the
SNP-exposure effect sizes against the SNP-outcome effect sizes, with
greater weight afforded to those SNPs with smaller SNP-outcome standard
errors, provides the estimate of the causal effect of the exposure on
the outcome. This is known as the inverse-variance weighted (IVW)
method.

@@

\subsubsection{Mendelian randomisation methods and their
assumptions}\label{mendelian-randomisation-methods-and-their-assumptions}

In this paper we consider three main classes of MR estimation. Full
details for each approach have been described extensively elsewhere.

@@

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
estimate the variance of the underlying liability explained by the SNP,
\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
\(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the log odds ratio and
\(p\) is the allele frequency of the SNP in the population (15). \(p\)
can be estimated using the allele frequency of the SNP in an ascertained
sample by deriving the \(2 \times 2\) contingency table from the odds
ratio \(e^\beta\), allele frequency in the ascertained sample
\$p\_\{cc\}, and number of cases \(N_1\) and controls \(N_0\).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total. In the context of this analysis, each method is
considered to be an `expert', taking a set of SNP-exposure and
SNP-outcome effect sizes and their standard errors as inputs. Our
objective is to select the expert most likely to be correct for a
specific MR analysis.

The mixture of experts method is a machine learning approach (16) which
seeks to divide a parameter space into subdomains, such that a
particular expert is used primarily for problems that reside in a
subdomain most suited to that expert. In this case our objective is to
identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations. This involves creating a `gating
function', whose purpose is to decide which expert to use for a specific
MR analysis, given the parameter space that is occupied by that dataset.
The metrics are a mixture of regression diagnostics and are described in
Supplementary table 2.

@@

For each simulation we used linear regression to estimate the genetic
effect of each SNP \(M\) on \(x\) in sample 1, and each SNP \(M\) on
\(y\) in sample 2. We then perform MR analysis in both directions,
retaining SNPs that have \(p < 5e-8\) in sample 1 to perform MR of \(x\)
on \(y\) (the true causal direction for non-null simulations), and
retaining SNPs that have \(p < 5e-8\) in sample 2 to perform MR of \(y\)
on \(x\) (the reverse causal direction for non-null simulations). We
treat the summary data (effect sizes and standard errors) used for
estimating \(x \rightarrow y\) the summary data used for estimating
\(y \rightarrow x\) as two separate datasets Hence, for each simulation
two datasets are generated which are analysed to produce 28 MR estimates
each. We performed 100,000 simulations using these parameters, resulting
in 200,000 datasets.

@@

We aim to maximise statistical power for datasets where
\(\beta_{x,y} \neq 0\) and minimise false discovery rates for datasets
where \(\beta_{x,y} = 0\). To train random forest decision trees to
predict performance for a particular method
\(h(O_{w,d}, \textbf{z}_{d}\) is generated where the training set of
input metrics for dataset \(d\) is \(\textbf{z}_{d}\) and the response
(optimisation function) is

@@

For each training dataset we record a set of 53 metrics
\(\textbf{z}_{d}\) (Supplementary table 2), and an outcome \(O_{w,d}\),
which is a measure of how well that method performed for each particular
simulated dataset. For each of our 28 methods, we need to create a model
that predicts the performance of the method based on metrics generated
from a dataset. To do this, for each method we train random forest
decision trees to predict that method's performance using the dataset's
metrics. The random forest approach is well suited to this problem
because there are likely to be non-continuous combinations of different
metrics that improve on prediction over, for example, a simple linear
model that does not learn about interactions.

@@

Figure 1b shows that the tophits strategy led to 49\% of the instruments
being primarily associated with either confounders or the outcome
phenotype, not the exposure phenotype. Applying Steiger filtering
reduced this to 25\%. Consequently, the false discovery rates (FDR) for
12 of the 14 methods reduced substantially when applied using Steiger
filtered instruments (Figure 1c). The true positive rates for the
methods based on Steiger filtering did however reduce slightly for 10 of
the 14 methods.

@@

Following evidence that the Steiger filtering approach can improve on
existing methods, we next hypothesised that a mixture of experts (MoE)
model would be able to use dataset characteristics to predict the most
appropriate of the 28 MR methods and instrument selection strategies
(Figure 2).

The ability to predict the performance for each of these methods is
shown in table 1, with prediction \(R^2\) ranging between 0.04 and 0.24.
The dataset characteristics with the most importance for each of the
predictors differed substantially between each dataset, as well as the
frequency for which each of the methods was selected in the testing
datasets. The FDR of each method when chosen, compared to their averages
across all datasets, reduced; and likewise the true positive rates of
each method when chosen increased compared to their simulation-wide
averages.

@@

The frequencies of the methods chosen by the MR-MoE analysis are shown
in Supplementary table 2. Amongst those deemed `significant', the IVW
fixed effects analysis using tophits instruments (the only method
applicable when there is no evidence of horizontal pleiotropy of any
sort) was selected in only 10.4\% of cases.

@@

Following a recent flurry in development of MR methodology, we have
devised a machine learning approach, MR-MoE, that seeks to predict the
performance of each MR method, selecting the one which is most likely to
maximise power whilst minimising FDR for a specific dataset. The mixture
of experts approach that we present here is trained using random forest
decision trees applied to extensive and diverse simulations, and we
demonstrate that it makes substantial performance improvements over any
other single method.

Crucially, we observed in our simulations that none of the available
methods were conservative and all, including MR-MoE, had FDR above 0.2.
Instrumental variables in Mendelian randomisation are typically chosen
blind, with GWAS significance being the only criteria. We illustrated
how confounders and reverse causal associations can easily lead to
invalid instruments being used in MR, and that even if those instruments
that only influence the exposure directly, most randomly generated
patterns of horizontal pleiotropy cannot be adequately accounted for by
any single MR method.

We applied the method to a curated set of GWAS summary data that was
present in MR-Base (22), generating hundreds of thousands of MR
estimates. MR-MoE selected a method that indicated some pattern of
horizontal pleiotropy for almost 90\% of cases, suggesting that it is
the rule rather than the exception. Based on our simulations we expect
that those associations that we reported to be `significant' are liable
to high type 1 error rates.

@@

A large proportion of the associations that were estimated used only a
single instrumental variable. Methods are emerging to attempt to
delineate between models of reverse cause, pleiotropy and multiple
causal variants in the same region (23,24), but separating vertical from
horizontal pleiotropy with a single instrument is not yet possible in
the two-sample MR framework. Other problems can also manifest, for
example frailty effects could induce associations for late-onset traits
(25). We also encounter a potential new problem when it comes to using
machine learning approaches to infer the correct model for a particular
dataset, namely, that the data is choosing the model. This is liable to
increase type 1 error rates, even though we have attempted to separate
the information used for optimisation from the information used to
predict performance. Though MR-MoE has high type 1 error rates, they
still remain amongst the lowest compared to all other methods that do
not suffer from this issue.

Despite these limitations, the construction of a causal graph of
`everything versus everything' does have appeal. First, though causality
is not guaranteed by MR, it can still be highly informative for
confirming or negating specific hypotheses, or to search for novel
putative associations. The potential to exploit GWAS summary data within
the properties of graph databases to aid with both of these endeavours
could be transformative in biological research.

\subsection{Figures}\label{figures}

@@

\begin{figure}
\centering
\includegraphics{images/fig1.pdf}
\caption{Simulations. a) Schematic of how GWAS with sufficient power can
lead to the selection of instrumental variables that are invalid. We
used arbitrary numbers of SNPs and confounders to simulate GWAS summary
datasets. b) Any SNP that has a direct influence on the exposure, or an
influence on a non-confounding intermediate variable, is considered a
`direct' effect. The bar-chart shows the number of direct effect
instruments retained for analysis using either the tophits approach or
the Steiger approach for instrument selection. c) Top: The false
discovery rates from null simulations for each of the 14 methods using
either tophits, Steiger filtered variants, or variants that are known to
be directly associated with the exposure (oracle, note that direct
effects can still exhibit horizontal pleiotropy in these simulations).
Bottom: The statistical power to detect true causal associations in the
non-null simulations.}
\end{figure}

@@

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Learning pleiotropic signatures to construct a causal map of the human phenome},
            pdfauthor={Gibran Hemani, Jie Zheng, Philip Haycock, Jack Bowden, Tom Gaunt, George Davey Smith},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Learning pleiotropic signatures to construct a causal map of the human
phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Gibran Hemani, Jie Zheng, Philip Haycock, Jack Bowden, Tom Gaunt, George
Davey Smith}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{03 July 2017}

@@

Mendelian randomization (MR) (1,2) exploits genetic pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy. Such a genetic variant is considered to be an instrumental
variable for the exposure, and can be exploited to mimic a randomised
controlled trial, enabling a causal estimate to be made by comparing the
outcome phenotypes between those individuals that have the
exposure-increasing allele against those who do not. Multiple
independent genetic variants for a particular exposure can be used
jointly to improve causal inference, because a) each variant represents
an independent natural experiment, and an overall causal estimate can be
obtained by meta-analysing the single estimates from each instrument;
and b) some violations of the assumptions of MR (most notably that the
SNP influences the outcome through no pathways other than the exposure)
can be evaluated through sensitivity analyses (3--7).

Genome-wide association studies (GWAS) have identified genetic
instrumental variables for thousands of phenotypes (8). Recent
developments in MR have enabled knowledge of instrumental variables to
be applied using only summary level data (known as two-sample MR, 2SMR)
(9). Here, in order to infer the causal effect of an exposure on an
outcome all that is required is an estimate of the genetic effects of
the instrumenting SNP on the exposure, and the corresponding estimate of
the effect on the outcome. This has two major advantages. First, GWAS
summary data is non-disclosive and often publicly available. Second,
causal inference can be made between phenotypes even if they have not
been measured in the same samples, limiting the breadth of possible
causal estimates only to the availability of GWAS summary data for the
traits in question (10).

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure. The
extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of HP (3--6). It is considered best
practice to report estimates from all available methods as sensitivity
analyses when presenting causal estimates, however this strategy is not
necessarily optimal for several reasons. First, if different methods
disagree it is not possible to know which is correct because the true
nature of HP exhibited by the instruments is not known. Second, though
the IVW approach is most statistically powerful under no HP, it can have
high false negative or low true positive rates in the presence of HP
compared to other methods. Given that pleiotropy has been hypothesised
to be universal (11,12), defaulting to the IVW method in the first
instance and using other methods as sensitivity analyses may not be
appropriate. Third, the available methods do not cover all possible
models of HP, and therefore an automated method for instrument selection
may be necessary. Fourth, it could be of interest to make causal effect
estimates for thousands of traits, in which case a critical evaluation
of each causal estimate of interest may not be possible or convenient.

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we introduce an approach to discard
genetic variants that are likely to be invalid. Second, we hypothesised
that characteristics of the summary data could indicate which method
would be most reliable, and we introduce new machine learning approaches
that attempt to automate both instrument and method selection. Using
curated GWAS summary data for thousands of phenotypes (10), we use these
new methods to construct a graph of millions of causal estimates. We
consider this to be a `draft' of the causal map of the human phenome,
but raise caution throughout that its interpretation is far from
straightforward

@@

The use of summary data in two-sample MR is described in detail
elsewhere (10). A brief outline of the procedure is as follows. First,
genetic instruments for the exposure trait need to be identified - those
SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
assumption of MR (that the instrument associates with the exposure) is
satisfied. We collect their effect sizes, standard errors and effect
alleles for the association with the exposure trait. These can be
obtanied manually from complete summary data or from curated lists of
significant GWAS associations. Next, the effects of those SNPs on the
outcome need to be obtained, typically necessitating complete summary
data because it is unlikely that these SNPs will have reached
genome-wide significance (and therefore be present in curated
catalogues). At its most simple implementation, the regression of the
SNP-exposure effect sizes against the SNP-outcome effect sizes, with
greater weight afforded to those SNPs with smaller SNP-outcome standard
errors, provides the estimate of the causal effect of the exposure on
the outcome. This is known as the inverse-variance weighted (IVW)
method.

@@

\subsubsection{MR methods and their
assumptions}\label{mr-methods-and-their-assumptions}

In this paper we consider three main classes of MR estimation. Full
details for each approach have been described previously. A summary of
the methods is given in Supplementary table 2.

@@

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
estimate the variance of the underlying liability explained by the SNP,
\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
\(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the log odds ratio and
\(p\) is the allele frequency of the SNP in the population (15). \(p\)
can be estimated using the allele frequency of the SNP in an ascertained
sample by deriving the \(2 \times 2\) contingency table from the odds
ratio \(e^\beta\), allele frequency in the ascertained sample
\$p\_\{cc\}, and number of cases \(N_1\) and controls \(N_0\). We
exclude any SNPs that do not have the correct causal direction for the
hypothesis, and do not have Steiger test \(p < 0.05\).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total (Supplementary table 2). In the context of this
analysis, each method is considered to be an `expert', taking a set of
SNP-exposure and SNP-outcome effect sizes and their standard errors as
inputs. Our objective is to select the expert most likely to be correct
for a specific MR analysis.

The mixture of experts method is a machine learning approach (16) which
seeks to divide a parameter space into subdomains, such that a
particular expert is used primarily for problems that reside in a
subdomain most suited to that expert. In this case our objective is to
identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations. This involves creating a `gating
function', whose purpose is to decide which expert to use for a specific
MR analysis, given the parameter space that is occupied by that dataset.
The metrics are a mixture of regression diagnostics and are described in
Supplementary table 3.

@@

For each simulation we used linear regression to estimate the genetic
effect of each SNP \(M\) on \(x\) in sample 1, and each SNP \(M\) on
\(y\) in sample 2. We then perform MR analysis in both directions,
mimicking GWAS by retaining SNPs that have \(p < 5e-8\) in sample 1 to
perform MR of \(x\) on \(y\) (the true causal direction for non-null
simulations), and retaining SNPs that have \(p < 5e-8\) in sample 2 to
perform MR of \(y\) on \(x\) (the reverse causal direction for non-null
simulations). We treat the summary data (effect sizes and standard
errors) used for estimating \(x \rightarrow y\) the summary data used
for estimating \(y \rightarrow x\) as two separate datasets Hence, for
each simulation two datasets are generated which are analysed to produce
28 MR estimates each. We performed 100,000 simulations using these
parameters, resulting in 200,000 datasets.

@@

We aim to maximise statistical power for datasets where
\(\beta_{x,y} \neq 0\) and minimise false discovery rates for datasets
where \(\beta_{x,y} = 0\). To train random forest decision trees to
predict performance for a particular method
\(h(O_{w,d}, \textbf{z}_{d})\) is generated where the training set of
input metrics for dataset \(d\) is \(\textbf{z}_{d}\) and the response
(optimisation function) is

@@

For each training dataset we record a set of 53 metrics
\(\textbf{z}_{d}\) (Supplementary table 3), and an outcome \(O_{w,d}\),
which is a measure of how well that method performed for each particular
simulated dataset. For each of our 28 methods, we need to create a model
that predicts the performance of the method based on metrics generated
from a dataset. To do this, for each method we train random forest
decision trees to predict that method's performance using the dataset's
metrics. The random forest approach is well suited to this problem
because there are likely to be non-continuous combinations of different
metrics that improve on prediction over, for example, a simple linear
model that does not learn about interactions.

@@

Figure 1b shows that the tophits strategy led to over half of the
instruments being primarily associated with either confounders or the
outcome phenotype, not the exposure phenotype. The proportion of invalid
instruments due to reverse cause increased as GWAS discovery sample size
increased. Applying Steiger filtering reduced this to 25\%.
Consequently, the false discovery rates (FDR) for 12 of the 14 methods
reduced substantially when applied using Steiger filtered instruments
(Figure 1c). The true positive rates for the methods based on Steiger
filtering did however reduce slightly for 10 of the 14 methods.

@@

Following evidence that the Steiger filtering approach can improve on
existing methods, we next hypothesised that a mixture of experts (MoE)
model would be able to predict the most appropriate of the 28 MR methods
and instrument selection strategies to apply to a particular dataset
based on its characteristics (Figure 2).

The ability to predict the performance for each of these methods is
shown in Supplementary table 2, with prediction \(R^2\) ranging between
0.04 and 0.24. The dataset characteristics with the most importance for
each of the predictors differed substantially between each dataset, as
well as the frequency for which each of the methods was selected in the
testing datasets. The FDR of each method when chosen, compared to their
averages across all datasets, reduced; and likewise the true positive
rates of each method when chosen increased compared to their
simulation-wide averages.

@@

The frequencies of the methods chosen by the MR-MoE analysis are shown
in Supplementary table 3. Amongst those deemed `significant', the IVW
fixed effects analysis using tophits instruments (the only method
applicable when there is no evidence of horizontal pleiotropy of any
sort) was selected in only 10.4\% of cases.

@@

The trend of increasing GWAS sample sizes continues, but while the
opportunity that this affords MR to be furnished with more instruments
is typically welcomed, here we have shown that it is invalid
instruments, exhibiting horizontal pleiotropy or reverse causation, that
are more likely to be identified than valid ones. Strategising how MR is
to be used in practice, therefore, must consider horizontal pleiotropy
to be the rule rather than the exception. Following a recent flurry in
development of MR methodology, we have devised a machine learning
approach, MR-MoE, that seeks to predict the performance of each MR
method, selecting the one which is most likely to maximise power whilst
minimising FDR for a specific dataset. The mixture of experts approach
that we present here is trained using random forest decision trees
applied to extensive and diverse simulations, and we demonstrate that it
makes substantial performance improvements over any other single method
in the presence or absence of extensive horizontal pleiotropy.

Crucially however, we observed in our simulations that none of the
available methods were conservative and all, including MR-MoE, had FDR
above 0.2. Instrumental variables in MR are typically chosen blind, with
GWAS significance being the only criteria. We illustrated how
confounders and reverse causal associations can easily lead to invalid
instruments being used in MR, and that even if those instruments that
only influence the exposure directly are used, most randomly generated
patterns of horizontal pleiotropy cannot be adequately accounted for by
any single MR method.

We applied the method to the curated set of GWAS summary data present in
MR-Base (22), generating several million MR estimates. MR-MoE selected a
method that indicated some pattern of horizontal pleiotropy for almost
90\% of cases, reinforcing the notion that it is the rule rather than
the exception. Based on our simulations we expect that those
associations that we reported to be `significant' are liable to high
type 1 error rates.

@@

A large proportion of the associations that were estimated used only a
single instrumental variable. Methods are emerging to attempt to
delineate between models of reverse cause, pleiotropy and multiple
causal variants in the same region (23,24), but separating vertical from
horizontal pleiotropy with a single instrument is not yet possible in
the two-sample MR framework. Other problems can also manifest, for
example frailty effects could induce associations for late-onset traits
(25); genetic variants could strongly relate to several phenotypes
making it difficult to ascertain which of them is the causal exposure
(26); and the measured feature for which genetic associations are known
could relate in complicated ways to the biological entity that are truly
causal (2).

We also encounter a potential new problem in using machine learning
approaches to infer the correct model for a particular dataset, namely,
that the data is choosing the model. This is liable to increase type 1
error rates, even though we have attempted to separate the information
used for optimisation from the information used to predict performance.
Though MR-MoE does exhibit higher type 1 error rates than are desirable,
they still remain amongst the lowest compared to all other MR methods
that do not suffer from this issue.

Despite these limitations, the construction of a causal graph of
`everything versus everything' does have appeal. First, though causality
is not guaranteed by MR, it can still be highly informative for
confirming or negating specific hypotheses. Second, it offers new
approaches to searching for novel putative associations. The potential
to exploit GWAS summary data within the properties of graph databases to
aid with both of these endeavours could be transformative in biological
research.

@@

\begin{figure}
\centering
\includegraphics{images/fig1.pdf}
\caption{Simulations. a) Schematic of how GWAS with sufficient power can
lead to the selection of instrumental variables that are invalid. We
used arbitrary numbers of SNPs and confounders to simulate GWAS summary
datasets. b) Any SNP that has a direct influence on the exposure, or an
influence on a non-confounding intermediate variable, is considered a
`direct' effect. The y-axis shows the proportion of instruments selected
for analysis due that are either direct associations with the exposure,
instruments for the outcome (reverse), or instruments for confounding
traits (confounder). The proportions are compared over a range of
different exposure discovery sample sizes (x-axis) and using either the
tophits approach (left) or the Steiger approach (right) for instrument
selection. c) Top: The false discovery rates from null simulations for
each of the 14 methods using either tophits, Steiger filtered variants,
or variants that are known to be directly associated with the exposure
(oracle, note that direct effects can still exhibit horizontal
pleiotropy in these simulations). Bottom: The statistical power to
detect true causal associations in the non-null simulations.}
\end{figure}

@@

\hypertarget{ref-Burgess2014a}{}
26. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
multivariable Mendelian randomization to disentangle the causal effects
of lipid fractions. PloS one {[}Internet{]}. 2014 Jan;9(10):e108891.
Available from:
\url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0108891}
